---
title: "Torch epiclock Training in R"
format: 
  html:
    theme: minty
    toc: true
    toc-expand: 3
    toc-location: left
    embed-resources: true
editor: visual
execute:
  warning: FALSE
---

### tl;dr

This script trains a Torch model for age prediction using DNA methylation data in R.

It includes data loading, preprocessing, model training, evaluation, and visualization.

The pipeline is structured to be reproducible and modular, allowing for easy adjustments to hyperparameters and probe sets on the VAI HPC with GPU settings.

### 1. setup

```{r}
# Load Libraries
library(torch)
library(data.table)
library(dplyr)
library(tidyr)
library(arrow)
library(rsample)
library(yardstick)
library(ggplot2)
library(patchwork)
library(reticulate)

# --- Configuration ---
# Set a project directory
project_dir <- "/varidata/research/projects/laird/jaemin.park/HPC/deeplearned-epiclock"
knitr::opts_knit$set(root.dir = project_dir)

# Set seed for reproducibility
seed <- 42
set.seed(seed)
torch::cuda_is_available()
torch_manual_seed(seed)

# --- Device and Hyperparameter Configuration ---
device <- if (cuda_is_available()) torch_device("cuda") else torch_device("cpu")
cat(paste("Using device:", device$type, "\n"))

# Training Hyperparameters
num_epochs     <- 100
batch_size     <- 128
learning_rate  <- 1e-4
weight_decay   <- 1e-5
patience       <- 15      # For early stopping
dropout_rate   <- 0.4
split_ratios   <- list(train = 0.70, val = 0.15, test = 0.15)
num_workers    <- 0       # Dataloader workers

# --- Data Paths ---
beta_matrix_path <- file.path(project_dir, "beta_matrix_dense.npy")
probe_ids_path   <- file.path(project_dir, "cpgprobes_from_mtx.npy")
sample_ids_path  <- file.path(project_dir, "gsmid_mtx_from_mtx.npy")
metadata_path    <- file.path(project_dir, "meta_filtered.feather")
probe_list_dir   <- file.path(project_dir, "scripts/probe_lists")

# --- Experiment Definition ---
# Define different sets of CpG probes to train models on
probe_sets_to_run <- list(
  "all_probes"     = NULL,
  "polycomb_CGI"   = file.path(probe_list_dir, "polycomb_CGI_probes.txt"),
  "PMDsoloWCGW"    = file.path(probe_list_dir, "PMDsoloWCGW_probes.txt"),
  "most_variable"  = file.path(probe_list_dir, "most_variable_probes.txt")
)

# Use reticulate to load Python's numpy arrays
np <- reticulate::import("numpy", convert = FALSE)
```

### 2. model architecture

```{r}
# Define the MLP Model using nn_module
age_predictor_mlp <- nn_module(
  "AgePredictorMLP",
  initialize = function(input_size,
                        hidden1_size = 512,
                        hidden2_size = 256,
                        dropout_rate = 0.4) {
    self$network <- nn_sequential(
      # Layer 1
      nn_linear(input_size, hidden1_size),
      nn_batch_norm1d(hidden1_size),
      nn_relu(),
      nn_dropout(dropout_rate),
      # Layer 2
      nn_linear(hidden1_size, hidden2_size),
      nn_batch_norm1d(hidden2_size),
      nn_relu(),
      nn_dropout(dropout_rate),
      # Output Layer
      nn_linear(hidden2_size, 1)
    )
  },
  forward = function(x) {
    self$network(x)
  }
)

# Display an example model structure
model_instance <- age_predictor_mlp(input_size = 10000)
print(model_instance)
```

### 3. Main Training Loop

```{r}
results_summary <- list()

# Iterate through each defined probe set
for (probe_set_name in names(probe_sets_to_run)) {
  
  cat(paste0("\n--- Starting Experiment: ", probe_set_name, " ---\n"))
  
  # 1. DATA LOADING AND PREPROCESSING
  cat("1. Loading and preprocessing data...\n")
  X_full         <- py_to_r(np$load(beta_matrix_path))
  all_probe_ids  <- as.character(py_to_r(np$load(probe_ids_path, allow_pickle=TRUE)))
  all_sample_ids <- toupper(trimws(as.character(py_to_r(np$load(sample_ids_path, allow_pickle=TRUE)))))
  meta_df        <- arrow::read_feather(metadata_path)
  
  # Align methylation data with metadata
  X_full         <- t(X_full)
  meta_dt        <- as.data.table(meta_df)[, .(gsm = toupper(trimws(as.character(gsm))), age_years)]
  samples_dt     <- data.table(gsm = all_sample_ids, original_order = 1:length(all_sample_ids))
  meta_aligned   <- meta_dt[samples_dt, on = "gsm", nomatch=0][order(original_order)]
  
  # Filter out samples with missing age
  valid_indices  <- which(!is.na(meta_aligned$age_years))
  X_full         <- X_full[valid_indices, , drop = FALSE]
  ages           <- as.numeric(meta_aligned$age_years[valid_indices])
  
  # Filter probes based on the current experiment's list
  probe_file_path <- probe_sets_to_run[[probe_set_name]]
  if (!is.null(probe_file_path)) {
    target_probes <- readLines(probe_file_path)
    probe_mask    <- all_probe_ids %in% target_probes
    X_filtered    <- X_full[, probe_mask, drop = FALSE]
  } else {
    X_filtered <- X_full
  }
  n_features <- ncol(X_filtered)
  cat(paste("  Prepared data with", n_features, "features.\n"))

  # 2. DATA SPLITTING & SCALING
  cat("2. Splitting and scaling data...\n")
  # Split data into training, validation, and test sets
  split_obj      <- initial_split(data.frame(idx = 1:nrow(X_filtered)), prop = split_ratios$train)
  train_indices  <- training(split_obj)$idx
  temp_indices   <- testing(split_obj)$idx
  val_test_split <- initial_split(data.frame(idx = temp_indices), prop = split_ratios$val / (split_ratios$val + split_ratios$test))
  val_indices    <- training(val_test_split)$idx
  test_indices   <- testing(val_test_split)$idx
  
  # Scale features based on the training set
  train_means    <- colMeans(X_filtered[train_indices, ], na.rm = TRUE)
  train_sds      <- apply(X_filtered[train_indices, ], 2, sd, na.rm = TRUE)
  train_sds[is.na(train_sds) | train_sds == 0] <- 1 # Avoid division by zero
  
  X_train_scaled <- scale(X_filtered[train_indices, ], center = train_means, scale = train_sds)
  X_val_scaled   <- scale(X_filtered[val_indices, ], center = train_means, scale = train_sds)
  X_test_scaled  <- scale(X_filtered[test_indices, ], center = train_means, scale = train_sds)

  y_train <- ages[train_indices]; y_val <- ages[val_indices]; y_test <- ages[test_indices]
  
  # 3. DATALOADER CREATION
  cat("3. Creating PyTorch dataloaders...\n")
  train_dataset <- tensor_dataset(
    torch_tensor(X_train_scaled, dtype = torch_float32()),
    torch_tensor(y_train, dtype = torch_float32())$unsqueeze(2)
  )
  val_dataset <- tensor_dataset(
    torch_tensor(X_val_scaled, dtype = torch_float32()),
    torch_tensor(y_val, dtype = torch_float32())$unsqueeze(2)
  )
  test_dataset <- tensor_dataset(
    torch_tensor(X_test_scaled, dtype = torch_float32()),
    torch_tensor(y_test, dtype = torch_float32())$unsqueeze(2)
  )
  
  train_loader <- dataloader(train_dataset, batch_size = batch_size, shuffle = TRUE, num_workers = num_workers)
  val_loader   <- dataloader(val_dataset, batch_size = batch_size, shuffle = FALSE, num_workers = num_workers)
  test_loader  <- dataloader(test_dataset, batch_size = batch_size, shuffle = FALSE, num_workers = num_workers)

  # 4. MODEL INITIALIZATION AND TRAINING
  cat("4. Initializing and training the model...\n")
  model <- age_predictor_mlp(input_size = n_features, dropout_rate = dropout_rate)$to(device = device)
  optimizer <- optim_adamw(model$parameters, lr = learning_rate, weight_decay = weight_decay)
  scheduler <- lr_reduce_on_plateau(optimizer, patience = 5, factor = 0.2)
  
  best_val_loss <- Inf
  epochs_no_improve <- 0
  history <- data.frame()

  for (epoch in 1:num_epochs) {
    # Training phase
    model$train()
    train_loss <- 0
    coro::loop(for (b in train_loader) {
      optimizer$zero_grad()
      outputs <- model(b[[1]]$to(device = device))
      loss <- nnf_smooth_l1_loss(outputs, b[[2]]$to(device = device))
      loss$backward()
      optimizer$step()
      train_loss <- train_loss + loss$item()
    })
    
    # Validation phase
    model$eval()
    val_loss <- 0
    all_preds <- c(); all_labels <- c()
    with_no_grad({
      coro::loop(for (b in val_loader) {
        outputs <- model(b[[1]]$to(device = device))
        loss <- nnf_smooth_l1_loss(outputs, b[[2]]$to(device = device))
        val_loss <- val_loss + loss$item()
        all_preds <- c(all_preds, as.numeric(outputs$cpu()))
        all_labels <- c(all_labels, as.numeric(b[[2]]$cpu()))
      })
    })
    
    # Log metrics and check for early stopping
    epoch_val_loss <- val_loss / length(val_loader)
    epoch_val_mae <- mae_vec(truth = all_labels, estimate = pmax(0, all_preds))
    history <- rbind(history, data.frame(epoch, train_loss = train_loss/length(train_loader), val_loss = epoch_val_loss, val_mae = epoch_val_mae))
    cat(sprintf("  Epoch %02d: Val Loss: %.4f, Val MAE: %.4f\n", epoch, epoch_val_loss, epoch_val_mae))
    
    scheduler$step(epoch_val_loss)
    if (epoch_val_loss < best_val_loss) {
      best_val_loss <- epoch_val_loss
      torch_save(model$state_dict(), "best_model.pt")
      epochs_no_improve <- 0
    } else {
      epochs_no_improve <- epochs_no_improve + 1
    }
    
    if (epochs_no_improve >= patience) {
      cat(paste("Early stopping at epoch", epoch, "\n"))
      break
    }
  }

  # 5. FINAL EVALUATION ON TEST SET
  cat("5. Evaluating on the test set...\n")
  # Load best model and evaluate
  best_model <- age_predictor_mlp(input_size = n_features, dropout_rate = dropout_rate)
  best_model$load_state_dict(torch_load("best_model.pt"))
  best_model$to(device = device)$eval()
  
  test_preds <- c()
  with_no_grad({
    coro::loop(for (b in test_loader) {
      outputs <- best_model(b[[1]]$to(device = device))
      test_preds <- c(test_preds, as.numeric(outputs$cpu()))
    })
  })
  test_preds_corr <- pmax(0, test_preds)
  
  final_mae <- mae_vec(truth = y_test, estimate = test_preds_corr)
  final_rsq <- rsq_vec(truth = y_test, estimate = test_preds_corr)
  final_ccc <- ccc_vec(truth = y_test, estimate = test_preds_corr)
  
  cat("  --- Test Set Performance ---\n")
  cat(sprintf("  MAE: %.4f\n  R²:  %.4f\n  CCC: %.4f\n", final_mae, final_rsq, final_ccc))
  
  # Store results
  results_summary[[probe_set_name]] <- list(
    ProbeSet = probe_set_name,
    Num_Features = n_features,
    MAE = final_mae,
    R2 = final_rsq,
    CCC = final_ccc,
    Epochs_Ran = epoch,
    History = history,
    Test_Predictions = data.frame(Actual = y_test, Predicted = test_preds_corr)
  )
}
```

### 4. Results & Visualization

```{r}
# Convert results list to a single data.table
results_df <- rbindlist(lapply(results_summary, function(x) {
  data.frame(ProbeSet=x$ProbeSet, Num_Features=x$Num_Features, MAE=x$MAE, R2=x$R2, CCC=x$CCC)
}))

# Print summary table
cat("\n--- Final Performance Summary ---\n")
print(results_df)

# --- Visualize Performance Comparison ---

# 1. MAE Bar Chart
p1 <- ggplot(results_df, aes(x = reorder(ProbeSet, MAE), y = MAE, fill = ProbeSet)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", MAE)), vjust = -0.5) +
  labs(title = "Mean Absolute Error (MAE) by Probe Set", x = "Probe Set", y = "MAE (Years)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 2. R-squared Bar Chart
p2 <- ggplot(results_df, aes(x = reorder(ProbeSet, -R2), y = R2, fill = ProbeSet)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.3f", R2)), vjust = -0.5) +
  labs(title = "R-squared (R²) by Probe Set", x = "Probe Set", y = "R-squared") +
  ylim(0, 1) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Display combined plot
print(p1 + p2)

# --- Visualize a Specific Result (e.g., the best one) ---
best_probe_set <- results_df$ProbeSet[which.min(results_df$MAE)]
cat(paste("\n--- Visualizing Best Model:", best_probe_set, "---\n"))

# 1. Training History Plot
history_data <- results_summary[[best_probe_set]]$History
p_hist <- ggplot(history_data, aes(x = epoch)) +
  geom_line(aes(y = train_loss, color = "Train Loss")) +
  geom_line(aes(y = val_loss, color = "Validation Loss")) +
  labs(title = paste("Training History for", best_probe_set), x = "Epoch", y = "Loss", color = "") +
  theme_minimal() + theme(legend.position="bottom")

# 2. Actual vs. Predicted Plot
pred_data <- results_summary[[best_probe_set]]$Test_Predictions
mae_val <- results_summary[[best_probe_set]]$MAE
r2_val <- results_summary[[best_probe_set]]$R2
p_pred <- ggplot(pred_data, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  coord_fixed() +
  labs(
    title = paste("Actual vs. Predicted Age for", best_probe_set),
    subtitle = sprintf("MAE = %.2f years, R² = %.3f", mae_val, r2_val),
    x = "Actual Age (Years)", y = "Predicted Age (Years)"
  ) +
  theme_minimal()

# Display plots for the best model
print(p_hist)
print(p_pred)
```
### sessionInfo

```{r}
sessionInfo()
```


